---
title: 'Matched-Tuple Randomization for Factorial Clinical
  Trial Designs: Theory and Application to Trial-Ready
  Cohorts'
author: 'Ronald G. Thomas, Ph.D.\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
abstract: |
  Factorial clinical trial designs efficiently evaluate
  multiple treatment factors simultaneously, but covariate
  balance across the resulting treatment combinations is
  difficult to achieve through simple randomization,
  particularly in moderate sample sizes. We propose
  matched-tuple randomization, in which subjects are grouped
  into sets of $k$ individuals with similar baseline
  covariate profiles -- where $k$ equals the number of
  factorial treatment combinations -- and each member of a
  tuple is assigned to a distinct treatment arm. We derive
  the randomization distribution, variance estimators, and
  an asymptotic relative-efficiency result versus simple
  randomization. A simulation study evaluates operating
  characteristics of the matched-tuple design itself (at two
  pool-oversampling ratios) under scenarios motivated by
  Alzheimer's disease prevention trials, where trial-ready
  cohorts provide large pools of characterized subjects
  amenable to matching; a head-to-head empirical comparison
  against simple, stratified, and rerandomization baselines
  is set up in the same simulation grid but had not yet been
  regenerated as of this draft (see the Provenance note in
  Results) and so is not yet reported. Software implementing
  the proposed methods is provided in the R package
  `factorialrandom`.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(tidyr)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  fig.align = 'center',
  results = 'asis',
  dev = 'pdf',
  cache = TRUE,
  cache.path = 'cache/'
)
options(
  knitr.kable.NA = '',
  knitr.table.format = 'latex',
  scipen = 1,
  digits = 3
)
```

# Introduction

## The Design Problem

Clinical trials increasingly evaluate combinations of
treatments. Anti-amyloid plus anti-tau therapies in
Alzheimer's disease, drug plus behavioral intervention in
oncology supportive care, and multi-component public health
programs all give rise to factorial treatment structures. A
$2 \times 2$ factorial design assigns subjects to one of four
combinations: neither factor, factor A alone, factor B alone,
or both. The hidden replication principle
[@fisherDesignExperiments1935] ensures that each subject
contributes to the estimation of both main effects and their
interaction, yielding the same precision as four separate
experiments of the same total size.

The challenge lies in covariate balance. With four or more
treatment arms, simple randomization can produce substantial
imbalance on prognostic covariates, inflating variance and
threatening the credibility of treatment comparisons. The
problem is compounded in moderate sample sizes ($n = 50$ to
$500$ per arm) typical of Phase II and prevention trials.

## Covariate Balance Strategies

Existing approaches to covariate balance -- stratified
randomization [@kernanStratifiedRandomization1999],
minimization [@pocockSequentialTreatment1975], and
rerandomization [@morganRerandomization2012] -- were developed
primarily for two-arm trials. Stratification is limited to a
few categorical covariates. Minimization extends to multiple
covariates but operates sequentially as subjects enroll,
which is unnecessary when the full pool is available at
design time. Rerandomization accepts only allocations meeting
a balance criterion but does not exploit the structure of
factorial contrasts.

Matched-pair randomization represents the most aggressive form
of restricted randomization. @baiOptimalityMatchedPairDesigns2022
proved that among all stratified designs with equal allocation
probabilities, a specific matched-pair design achieves minimum
asymptotic variance. @imaiVarianceIdentification2008 and
@fogartyRegressionAssisted2018 developed the inferential
framework. However, these results are restricted to $k = 2$
arms.

## Trial-Ready Cohorts

The trial-ready cohort (TRC) paradigm, developed for
Alzheimer's disease prevention research, pre-screens and
characterizes large pools of potential trial participants.
The TRC-PAD project [@aisenTRCPADOverview2020;
@aisenTRCPADExperience2020] enrolled over 30,000 individuals
in its web-based screening component and characterized
approximately 2,000 in-person participants with cognitive
testing, APOE genotyping, and amyloid biomarkers. The
AHEAD 3-45 study [@rafiiAHEAD2023] and DIAN-TU platform
[@batemanDIAN2012; @bateman2017DIANTU] draw on similar
infrastructure.

These characterized pools create an opportunity: when the
full cohort is available before randomization, multivariate
matching can be performed optimally rather than sequentially.
The rich baseline phenotyping (amyloid PET, tau PET, cognitive
composites, genetics) provides strong prognostic covariates,
and the factorial treatment structures emerging in combination
therapy trials create a natural demand for $k > 2$ matching.

## Contribution

We propose matched-tuple randomization for factorial designs.
Subjects are grouped into tuples of size $k$ (where $k$ is
the number of treatment combinations) based on multivariate
baseline covariates, and treatment assignment is constrained
so that each tuple member receives a distinct combination. We
develop:

1. Algorithms for forming matched $k$-tuples using
   Mahalanobis and prognostic score distances.
2. Exact and asymptotic variance formulas for factorial
   effects under the matched-tuple randomization distribution.
3. A simulation study evaluating efficiency gains under
   scenarios motivated by Alzheimer's disease prevention
   trials.
4. An R package implementing the methods.

# Background

## Factorial Designs

### Classical Framework

@fisherDesignExperiments1935 introduced factorial experiments
to evaluate multiple factors simultaneously. Consider $F$
factors, each at two levels ($+$ and $-$). The $2^F$ factorial
assigns subjects to all level combinations. Main effects and
interactions are defined as signed averages of cell means. The
key efficiency result is that every observation contributes to
every estimable effect.

@dasguptaCausalInference2015 recast factorial designs within
the potential outcomes framework. For a $2^2$ factorial, each
subject has four potential outcomes
$Y_i(00), Y_i(10), Y_i(01), Y_i(11)$ corresponding to the
four treatment combinations. The main effect of factor A is
defined as:
\begin{equation}
  \tau_A = \frac{1}{N} \sum_{i=1}^{N}
    \left[ \frac{Y_i(10) + Y_i(11)}{2} -
           \frac{Y_i(00) + Y_i(01)}{2} \right]
\end{equation}
and analogously for factor B and the interaction $AB$.

### Factorial Designs in Clinical Trials

Factorial designs are increasingly used in clinical trials
to evaluate combination therapies and multi-component
interventions. @collinsMultiphaseOptimization2007 introduced
the Multiphase Optimization Strategy (MOST), using factorial
experiments to identify active intervention components.

In Alzheimer's disease, the next generation of prevention
trials is expected to combine mechanisms of action: anti-amyloid
antibodies with anti-tau agents, secretase inhibitors with
lifestyle interventions, or multiple doses of the same agent
with and without an adjunct therapy
[@cummingsADTrialDesign2024]. These combinations create
natural factorial structures.

## Matched Randomization

### Matched-Pair Designs ($k = 2$)

In a matched-pair design, $N$ subjects are grouped into $M =
N/2$ pairs based on baseline covariates, and within each pair,
one subject is randomized to treatment and the other to
control. The treatment effect estimator is:
\begin{equation}
  \hat{\tau} = \frac{1}{M} \sum_{m=1}^{M}
    (Y_{m,T} - Y_{m,C})
\end{equation}
where $Y_{m,T}$ and $Y_{m,C}$ are the outcomes for the treated
and control subjects in pair $m$.

Under the matched-pair randomization distribution, the variance
of $\hat{\tau}$ is:
\begin{equation}
  \text{Var}(\hat{\tau}) = \frac{1}{M^2} \sum_{m=1}^{M}
    \frac{(\tau_m - \bar{\tau})^2 +
          \text{Var}_m}{2}
\end{equation}
where $\tau_m = Y_{m}(1) - Y_{m}(0)$ is the within-pair
treatment effect and $\text{Var}_m$ captures within-pair
variability in potential outcomes.

When pairs are well matched on prognostic covariates, the
within-pair treatment effects $\tau_m$ are similar to each
other, and the first term is small. This is the source of
precision gains.

### Optimal Matching

@greevyOptimalMatching2004 formulated optimal multivariate
matching as a minimum-cost network flow problem. For $k = 2$,
the problem reduces to minimum-weight perfect matching on a
bipartite graph, solvable in $O(N^3)$ time. The distance
between subjects $i$ and $j$ is typically the Mahalanobis
distance:
\begin{equation}
  d(i,j) = \sqrt{(X_i - X_j)' S^{-1} (X_i - X_j)}
\end{equation}
where $X_i$ is the covariate vector and $S$ is the sample
covariance matrix.

### Rerandomization

@morganRerandomization2012 proposed rerandomization as an
intermediate approach between simple randomization and
deterministic matching. A randomization is accepted only if
the Mahalanobis distance between treatment group covariate
means falls below a threshold $a$:
\begin{equation}
  M = (\bar{X}_T - \bar{X}_C)' \hat{\text{cov}}(\bar{X}_T -
    \bar{X}_C)^{-1} (\bar{X}_T - \bar{X}_C) \le a
\end{equation}

@liRerandomizationAsymptotics2018 derived the asymptotic
distribution of treatment effect estimators under
rerandomization, showing it is a truncated normal with
reduced variance.

## Trial-Ready Cohorts in Alzheimer's Disease

### The Recruitment Challenge

Alzheimer's disease prevention trials face screening failure
rates exceeding 80%, primarily because eligibility requires
confirmation of brain amyloid pathology via PET imaging or
cerebrospinal fluid analysis
[@sperlingTrialDesignAD2011]. The cost of screening a single
participant can exceed \$10,000. These constraints motivate
the trial-ready cohort approach, in which screening and
characterization are performed prospectively, creating a pool
of eligible participants ready for rapid enrollment.

### TRC-PAD

The Trial-Ready Cohort for Preclinical/Prodromal Alzheimer's
Disease (TRC-PAD) was established by the Alzheimer's Clinical
Trials Consortium [@aisenTRCPADOverview2020]. The architecture
includes:

- An online longitudinal screening study (APT Webstudy) with
  quarterly cognitive assessments. Over 30,000 individuals
  consented by 2020.
- Machine learning risk prediction using demographic, genetic,
  and cognitive data to identify individuals likely to have
  elevated brain amyloid.
- In-person screening with cognitive battery, APOE genotyping,
  and plasma biomarkers (p-tau217, A$\beta$42/40 ratio)
  [@mcdonoughTRCScreening2024].
- Longitudinal follow-up of biomarker-confirmed eligible
  participants until a matching trial opens.

### AHEAD 3-45

The AHEAD 3-45 study [@rafiiAHEAD2023] tests lecanemab in
preclinical AD using a sister-trial design: A3 (intermediate
amyloid, Phase 2, imaging endpoints) and A45 (elevated
amyloid, Phase 3, cognitive endpoints). Both share a common
screening process drawing from TRC-PAD and direct
recruitment. The study demonstrates how TRC infrastructure
feeds directly into clinical trials.

### DIAN-TU and EPAD

The Dominantly Inherited Alzheimer Network
[@batemanDIAN2012] maintains a registry of autosomal dominant
AD mutation carriers, providing a genetically defined
trial-ready cohort. The DIAN-TU platform [@bateman2017DIANTU]
implemented adaptive designs with shared placebo arms and
cognitive run-in periods.

The European Prevention of Alzheimer's Dementia (EPAD)
programme [@solomonEPAD2022] recruited over 2,000
participants into a longitudinal cohort across 39 European
institutions, explicitly designed as a readiness cohort for
adaptive proof-of-concept trials.

### Relevance to Matched Factorial Designs

Trial-ready cohorts provide the essential prerequisites for
matched-tuple randomization: large characterized pools with
rich baseline phenotyping. The pool-to-sample ratio (available
subjects divided by trial enrollment target) determines
matching quality. TRC-PAD's pool of thousands of characterized
individuals, combined with the factorial treatment structures
emerging in combination therapy trials, creates an ideal
application setting.

# Methods

## Notation

Consider $N$ subjects to be randomized across a $2^F$
factorial design with $k = 2^F$ treatment combinations.
Subjects are grouped into $M = N/k$ matched tuples, each
containing $k$ subjects with similar baseline covariate
profiles. Within each tuple, the $k$ treatment combinations
are assigned uniformly at random to the $k$ subjects.

Let $X_i \in \mathbb{R}^p$ denote the baseline covariate
vector for subject $i$. Let $Y_i(z)$ denote the potential
outcome for subject $i$ under treatment combination $z \in
\{0, 1\}^F$. The observed outcome is $Y_i = Y_i(Z_i)$ where
$Z_i$ is the assigned treatment.

## Matching Algorithm

### Distance Metric

We consider two distance metrics for tuple formation:

**Mahalanobis distance.** For subjects $i$ and $j$:
\begin{equation}
  d_M(i,j) = \sqrt{(X_i - X_j)' S^{-1} (X_i - X_j)}
\end{equation}
where $S$ is the pooled sample covariance matrix of the
covariates.

**Prognostic score distance.** When historical data or pilot
study results are available, fit a model $\hat{m}(X) =
E[Y(0) \mid X]$ predicting the outcome under the reference
condition. Then:
\begin{equation}
  d_P(i,j) = |\hat{m}(X_i) - \hat{m}(X_j)|
\end{equation}
Matching on the prognostic score concentrates precision gains
on the most relevant covariate dimension.

### Tuple Formation

Given $N$ subjects and tuple size $k$, partition the subjects
into $M = \lfloor N/k \rfloor$ tuples to minimize total
within-tuple dispersion:
\begin{equation}
  \min_{\mathcal{P}} \sum_{m=1}^{M}
    \sum_{i < j \in \mathcal{T}_m} d(i,j)
\end{equation}
where $\mathcal{P} = \{\mathcal{T}_1, \ldots,
\mathcal{T}_M\}$ is a partition into tuples.

For $k = 2$, this is optimal pair matching, solvable in
polynomial time. For $k > 2$ the exact partition problem is
combinatorially hard, and the `factorialrandom` package
currently implements one heuristic:

1. **Sequential greedy.** Select a seed subject, find the
   $k-1$ nearest neighbors, form a tuple, remove from pool,
   repeat. Order seeds by distance to the pool centroid
   (outermost first) to avoid stranding outliers.
   Implemented as `match_tuples(..., method = "greedy")`.

Hierarchical clustering and integer-programming formulations
of the same partition problem are natural extensions -- and
were part of the original design plan (see
`docs/analysis-plan.md`, Aim 1) -- but are not implemented in
the current package: `match_tuples(..., method = "optimal")`
currently aliases to the greedy heuristic rather than solving
an integer program, and no hierarchical-clustering code path
exists. All simulation results and package behavior described
below reflect the greedy heuristic only; benchmarking
alternative tuple-formation algorithms is left to future
work.

### Pool Oversampling

When the trial-ready cohort provides $N_{\text{pool}} > N$
characterized subjects, matching quality improves. The
algorithm selects the $N$ subjects that form the best $M$
tuples, leaving $N_{\text{pool}} - N$ subjects unmatched. The
pool-to-sample ratio $r = N_{\text{pool}} / N$ is a design
parameter: larger $r$ yields tighter within-tuple matching at
the cost of excluding more subjects.

## Randomization Distribution

Within each tuple $m$, the $k$ treatment combinations are
assigned to the $k$ subjects by drawing uniformly from the
$k!$ permutations. Tuples are independent, so the
randomization space has $(k!)^M$ equally likely allocations.

This constrained randomization guarantees that for every
tuple, each treatment arm contains exactly one subject.
Consequently, all factorial contrasts (main effects,
interactions) are perfectly balanced within each tuple.

## Estimation

### Difference-in-Means

For a factorial contrast $c$ defined by a coefficient vector
$w_z$ over treatment combinations (with $\sum_z w_z = 0$),
the estimator is:
\begin{equation}
  \hat{\tau}_c = \frac{1}{M} \sum_{m=1}^{M}
    \sum_{z} w_z \, Y_{m}(z)
\end{equation}
where $Y_{m}(z)$ is the outcome of the subject in tuple $m$
assigned to treatment $z$. This estimator is unbiased under
the matched-tuple randomization distribution.

### Variance Estimation

The exact variance under matched-tuple randomization is:
\begin{equation}
  \text{Var}(\hat{\tau}_c) = \frac{1}{M^2}
    \sum_{m=1}^{M} V_m(c)
\end{equation}
where $V_m(c)$ depends on the within-tuple potential
outcomes. A conservative estimator replaces $V_m(c)$ with
the within-tuple sample variance of the contrast-weighted
outcomes.

### Regression Adjustment

Following @linAgnosticNotes2013 and
@fogartyRegressionAssisted2018, we augment the
difference-in-means with a regression adjustment:
\begin{equation}
  \hat{\tau}_c^{\text{adj}} = \hat{\tau}_c -
    \hat{\beta}'(\bar{X}_c^{+} - \bar{X}_c^{-})
\end{equation}
where $\hat{\beta}$ is estimated from an outcome regression
and $\bar{X}_c^{+}, \bar{X}_c^{-}$ are the covariate means
for the positive and negative arms of contrast $c$. Under
matched-tuple randomization, $\bar{X}_c^{+} - \bar{X}_c^{-}$
is already close to zero, so the adjustment is small but
further reduces variance.

## Efficiency Analysis

### Relative Efficiency

Define the relative efficiency of matched-tuple randomization
versus simple randomization as:
\begin{equation}
  \text{RE} = \frac{\text{Var}_{\text{simple}}(\hat{\tau}_c)}
    {\text{Var}_{\text{matched}}(\hat{\tau}_c)}
\end{equation}

We derive RE as a function of the within-tuple intraclass
correlation $\rho_w$ of the contrast-relevant potential
outcomes. When tuples are perfectly matched
($\rho_w \to 1$), $\text{RE} \to k/(k-1) \cdot M$,
reflecting the elimination of between-tuple variation.

### Degrees of Freedom

A practical concern is the loss of degrees of freedom. Under
simple randomization with $N$ subjects and $k$ arms, the
residual degrees of freedom for variance estimation are
$N - k$. Under matched-tuple randomization, the effective
degrees of freedom are $M - 1 = N/k - 1$. When $k$ is large
and $M$ is small, this loss can offset precision gains from
matching, particularly for coverage of confidence intervals
[@martinMatchingPower1993, pp. 329-338]. The rule-of-thumb
threshold ($M > 30$) used in the Discussion is offered as a
heuristic consistent with the general degrees-of-freedom
argument in that source, not a value derived or directly
quoted from it; no page-specific numeric threshold from
@martinMatchingPower1993 is reproduced here.

# Simulation Study

## Design

We evaluate five randomization strategies in a $2 \times 2$
factorial design ($k = 4$) under scenarios motivated by
Alzheimer's disease prevention trials.

### Data Generating Mechanism

```{r dgm-params}
dgm_params <- data.frame(
  Factor = c(
    'Total sample size ($N$)',
    'Covariate dimension ($p$)',
    'Prognostic strength ($R^2$)',
    'Main effects A and B (SD units, equal)',
    'Interaction AB (SD units)',
    'Pool/sample ratio ($r$), matched methods only',
    'Outcome distribution'
  ),
  Levels = c(
    '100, 200, 500',
    '3, 8',
    '0.1, 0.3, 0.5',
    '0, 0.2',
    '0, 0.15',
    '1.0, 2.0',
    'Normal'
  )
)
kable(
  dgm_params,
  booktabs = TRUE,
  escape = FALSE,
  caption = 'Data generating mechanism parameters.'
) |>
  kable_styling(
    latex_options = c('hold_position'),
    font_size = 10
  )
```

Subjects have $p$ baseline covariates drawn from a
multivariate normal distribution with AR(1) correlation
($\rho = 0.3$). The outcome is generated as:
\begin{equation}
  Y_i = X_i' \beta + \tau_A \cdot A_i + \tau_B \cdot B_i +
    \tau_{AB} \cdot A_i B_i + \varepsilon_i
\end{equation}
where $\beta$ is calibrated to achieve the specified
$R^2$, $A_i$ and $B_i$ are treatment indicators, and
$\varepsilon_i \sim N(0, \sigma^2)$.

### Comparator Methods

```{r comparators}
comps <- data.frame(
  Method = c(
    'Simple randomization',
    'Stratified randomization',
    'Rerandomization (Mahalanobis)',
    'Matched-tuple, pool ratio 1.0 (proposed)',
    'Matched-tuple, pool ratio 2.0 (proposed)'
  ),
  Description = c(
    'Equal probability, no covariate balancing',
    'Stratify on top-2-covariate quartiles, permuted blocks',
    paste0('Accept allocations with ',
      '$M < \\chi^2_{p,0.001}$'),
    'Greedy $k$-tuples, random within-tuple, no oversampling',
    'Greedy $k$-tuples from a 2x pool, best tuples retained'
  )
)
kable(
  comps,
  booktabs = TRUE,
  escape = FALSE,
  caption = 'Randomization strategies compared.'
) |>
  kable_styling(
    latex_options = c('hold_position'),
    font_size = 10
  )
```

`docs/analysis-plan.md` originally listed minimization
(Pocock-Simon dynamic balancing) as a fifth comparator.
Minimization is omitted from the simulation because it is a
sequential-enrollment procedure: it balances covariates as
subjects arrive one at a time, which is not the setting this
paper addresses (a fully characterized pool available before
randomization). Rerandomization and matched-tuple
randomization are both pool-based comparators and so provide
a more direct contrast with the proposed method; minimization
is discussed as background in `docs/literature-review.md`
but is not implemented in `factorialrandom` and not simulated
here.

### Evaluation Metrics

For each scenario and method, we compute across 2,000
simulation replicates:

- Covariate balance: maximum absolute standardized mean
  difference across all pairwise arm comparisons.
- Bias of the main effect and interaction estimators.
- Variance ratio (relative to simple randomization).
- 95\% confidence interval coverage.
- Power (proportion of replicates rejecting $H_0: \tau = 0$
  at $\alpha = 0.05$).

## Results

### Headline findings

```{r headline-16, include=FALSE}
perf_path <- file.path(
  here::here(), "analysis", "data", "derived_data",
  "sim_performance.rds"
)
perf <- if (file.exists(perf_path)) readRDS(perf_path) else NULL

if (!is.null(perf)) {
  abs_bias_max <- max(abs(perf$bias), na.rm = TRUE)
  cov_range <- range(perf$coverage, na.rm = TRUE)
  rej_range <- range(perf$rejection_rate, na.rm = TRUE)
  n_scen_done <- nrow(perf)
  methods_done <- sort(unique(perf$method))
  n_methods_done <- length(methods_done)
} else {
  abs_bias_max <- NA_real_
  cov_range <- c(NA_real_, NA_real_)
  rej_range <- c(NA_real_, NA_real_)
  n_scen_done <- 0
  methods_done <- character(0)
  n_methods_done <- 0
}
```

**Provenance.** *As of this draft, `sim_performance.rds` was
produced before the `tuple_balance()` and
`regression_adjustment()` bug fixes described below were
applied, and covers only
`r n_methods_done` of the 5 planned randomization methods
(`r paste(methods_done, collapse = ", ")`).* Root cause: prior
to the fix, `tuple_balance()` crashed on `tuples = NULL`
(the value returned by `generate_factorial_trial()` for
every non-matched method), and `run_condition()` wrapped
each replicate in a `tryCatch()` that discarded the error, so
every replicate for `simple`, `stratified`, and
`rerandomization` failed silently across all scenarios. That
crash is now fixed (`tuple_balance()` treats `tuples = NULL`
as an unmatched design and computes raw covariate SMDs
across `treatment` directly), and `run_condition()` now
reports a warning with an error count instead of failing
silently. A reduced-scale check (4 scenarios, 20 replicates,
all 5 methods) after the fix produced non-NA bias, coverage,
and rejection-rate estimates for every method; see
`docs/pub_review_remediation_2026-08-16.md` for that
verification run.
Within the `r n_scen_done` scenario-method-estimator rows
that are currently on disk, absolute bias never exceeded
`r if (is.finite(abs_bias_max)) sprintf('%.3f', abs_bias_max) else 'NA'`,
coverage of 95% CIs fell in
[`r if (all(is.finite(cov_range))) sprintf('%.3f', cov_range[1]) else 'NA'`,
`r if (all(is.finite(cov_range))) sprintf('%.3f', cov_range[2]) else 'NA'`],
and rejection rate in
[`r if (all(is.finite(rej_range))) sprintf('%.3f', rej_range[1]) else 'NA'`,
`r if (all(is.finite(rej_range))) sprintf('%.3f', rej_range[2]) else 'NA'`],
but these ranges characterize matched-tuple designs only
(`matched_1.0`/`matched_2.0`), not a comparison against
simple, stratified, or rerandomization baselines,
and every figure and table below inherits that limitation
until the sweep is regenerated.
**TODO (before submission):** re-run
`Rscript analysis/scripts/01_run_simulation.R` in full (all 5
methods x 72 scenarios x 2,000 replicates) with the code
fixes in place, then regenerate this section and every
downstream figure/table from the new
`sim_performance.rds`. A full-scale rerun was not attempted
during this remediation pass: a reduced-scale timing check
(4 scenarios x 5 methods x 20 replicates, single core) took
about 13 seconds, which extrapolates to several hours for the
full 72 x 5 x 2,000 grid on a single core; running the full
sweep, or parallelizing `run_condition()` calls across
scenarios/methods (the script already loads the `parallel`
package but does not currently use it), is left to the user.

### Covariate Balance

```{r balance-asmd, fig.cap='Mean of the maximum absolute standardized mean difference across covariates by randomization method, prognostic strength ($R^2$), and sample size, for the matched-tuple methods currently in `sim\\_performance.rds` (matched\\_1.0, matched\\_2.0; see the Provenance note above).'}
if (!is.null(perf)) {
  gd <- perf |>
    dplyr::filter(effect == "F1") |>
    dplyr::group_by(method, n, r_squared) |>
    dplyr::summarise(asmd = mean(mean_max_asmd, na.rm = TRUE),
                     .groups = "drop")
  ggplot(gd, aes(x = factor(n), y = asmd,
                 fill = factor(r_squared))) +
    geom_col(position = "dodge") +
    facet_wrap(~ method, nrow = 1) +
    scale_fill_brewer(palette = "Blues", name = expression(R^2)) +
    labs(x = "Sample size per arm",
         y = "Mean max ASMD across covariates") +
    theme_bw(base_size = 9) +
    theme(legend.position = "bottom")
} else {
  cat("*Awaiting `sim_performance.rds`.*")
}
```

### Efficiency Gains

```{r efficiency-emp-se, fig.cap='Empirical SE of the main-effect estimator by prognostic strength ($R^2$) and sample size, separately for the three variance estimators (Neyman, paired, regression-adjusted). Data predate the \\texttt{regression\\_adjustment()} point-estimate fix described in the text; the adjusted and Neyman point estimates in this run were identical by construction, so this panel should be regenerated from the corrected simulation before being used to support any claim about the adjusted estimator.'}
if (!is.null(perf)) {
  gd <- perf |>
    dplyr::filter(effect == "F1", method == "matched_2.0") |>
    dplyr::group_by(n, r_squared, estimator) |>
    dplyr::summarise(emp_se = mean(emp_se, na.rm = TRUE),
                     .groups = "drop")
  ggplot(gd, aes(x = r_squared, y = emp_se,
                 colour = estimator, shape = estimator,
                 group = estimator)) +
    geom_line() +
    geom_point(size = 1.8) +
    facet_wrap(~ paste0("n = ", n), nrow = 1) +
    scale_colour_brewer(palette = "Set1") +
    labs(x = expression(R^2),
         y = "Empirical SE",
         colour = "Estimator", shape = "Estimator") +
    theme_bw(base_size = 9) +
    theme(legend.position = "bottom")
} else {
  cat("*Awaiting `sim_performance.rds`.*")
}
```

### Coverage and Power

```{r coverage-by-method-n, fig.cap='Empirical 95\\% CI coverage by randomization method, sample size, and variance estimator. The horizontal line marks the nominal 0.95.'}
if (!is.null(perf)) {
  gd <- perf |>
    dplyr::filter(effect == "F1") |>
    dplyr::group_by(method, n, estimator) |>
    dplyr::summarise(coverage = mean(coverage, na.rm = TRUE),
                     .groups = "drop")
  ggplot(gd, aes(x = factor(n), y = coverage,
                 colour = estimator, shape = estimator,
                 group = estimator)) +
    geom_hline(yintercept = 0.95, linetype = "dashed",
               colour = "grey50") +
    geom_line() +
    geom_point(size = 1.8) +
    facet_wrap(~ method, nrow = 1) +
    scale_colour_brewer(palette = "Set1") +
    coord_cartesian(ylim = c(0.85, 1.0)) +
    labs(x = "Sample size per arm",
         y = "Empirical 95% coverage",
         colour = "Estimator", shape = "Estimator") +
    theme_bw(base_size = 9) +
    theme(legend.position = "bottom")
} else {
  cat("*Awaiting `sim_performance.rds`.*")
}
```

**Monte Carlo standard error.** Table
\@ref(tab:mcse-table) reports the Monte Carlo standard errors
(MCSE) of the bias, coverage, and rejection-rate estimates
above, computed per Morris, White, and Crowther
[-@morris2019using, Table 6] as
$\text{MCSE}(\text{bias}) = \text{emp\_SE}/\sqrt{n_{\text{reps}}}$
and
$\text{MCSE}(p) = \sqrt{p(1-p)/n_{\text{reps}}}$ for the
coverage and rejection proportions.

```{r mcse-table}
if (!is.null(perf)) {
  mcse_tab <- perf |>
    dplyr::filter(effect == "F1", estimator == "neyman") |>
    dplyr::group_by(method) |>
    dplyr::summarise(
      mean_mcse_bias = mean(mcse_bias, na.rm = TRUE),
      mean_mcse_coverage = mean(mcse_coverage, na.rm = TRUE),
      mean_mcse_rejection = mean(mcse_rejection, na.rm = TRUE),
      .groups = "drop"
    )
  names(mcse_tab) <- c("Method", "MCSE(bias)",
                        "MCSE(coverage)", "MCSE(rejection)")
  kable(
    mcse_tab,
    booktabs = TRUE,
    digits = 4,
    caption = paste0(
      "Mean Monte Carlo standard errors of bias, coverage, ",
      "and rejection-rate estimates (main effect F1, Neyman ",
      "estimator), averaged over scenarios currently on disk."
    )
  ) |>
    kable_styling(latex_options = c("hold_position"),
                 font_size = 10)
} else {
  cat("*Awaiting `sim_performance.rds`.*")
}
```

### Effect of Pool Oversampling

```{r oversampling-relative-efficiency, fig.cap='Relative efficiency from 2x pool oversampling: ratio of empirical SE under matched\\_1.0 (no oversampling) to matched\\_2.0 (2x pool), by prognostic strength ($R^2$) and sample size. Values above 1 indicate that oversampling improves precision.'}
if (!is.null(perf)) {
  gd <- perf |>
    dplyr::filter(effect == "F1") |>
    dplyr::group_by(n, r_squared, estimator, method) |>
    dplyr::summarise(emp_se = mean(emp_se, na.rm = TRUE),
                     .groups = "drop") |>
    tidyr::pivot_wider(names_from = method,
                       values_from = emp_se) |>
    dplyr::mutate(rel_efficiency = matched_1.0 / matched_2.0)
  ggplot(gd, aes(x = r_squared, y = rel_efficiency,
                 colour = estimator, shape = estimator,
                 group = estimator)) +
    geom_hline(yintercept = 1, linetype = "dashed",
               colour = "grey50") +
    geom_line() +
    geom_point(size = 1.8) +
    facet_wrap(~ paste0("n = ", n), nrow = 1) +
    scale_colour_brewer(palette = "Set1") +
    labs(x = expression(R^2),
         y = "Empirical SE ratio (1.0 / 2.0)",
         colour = "Estimator", shape = "Estimator") +
    theme_bw(base_size = 9) +
    theme(legend.position = "bottom")
} else {
  cat("*Awaiting `sim_performance.rds`.*")
}
```

# Application: Alzheimer's Disease Prevention Trial

We illustrate the proposed method using a design scenario
motivated by the AHEAD trial structure. Consider a $2 \times
2$ factorial testing anti-amyloid therapy (yes/no) crossed
with a tau-targeting intervention (yes/no) in preclinical AD
subjects drawn from a trial-ready cohort.

## Design Parameters

- **Pool size:** 2,000 characterized TRC participants
- **Trial enrollment:** 500 (125 per arm)
- **Pool/sample ratio:** 4.0
- **Matching covariates:** amyloid PET (Centiloids), tau PET
  (SUVR), APOE4 carrier status, baseline PACC score, age
- **Primary endpoint:** Change in PACC at 4 years
- **Simulated prognostic strength, effects (design choice):**
  $R^2 = 0.3$, main effects $0.2$ SD each, interaction $0.15$
  SD. These are illustrative values chosen to sit in the
  middle of the main simulation grid (Table
  \@ref(tab:dgm-params)); they are not fitted to any specific
  AHEAD-3-45 pilot data.

```{r application-data, include=FALSE}
app_path <- file.path(
  here::here(), "analysis", "data", "derived_data",
  "application_example.rds"
)
app_perf <- if (file.exists(app_path)) readRDS(app_path) else NULL
```

The tables below are generated from a design-specific
simulation at the parameters listed above ($N_{\text{pool}} =
2{,}000$, $N = 500$, $r = 4.0$, $k = 4$, prognostic $R^2 =
0.3$, main effects $0.2$ SD, interaction $0.15$ SD), run in
`analysis/scripts/02_application_example.R`. This run uses
`r if (!is.null(app_perf)) unique(app_perf$n_reps) else "NA"`
replicates per method rather than the 2,000 used in the main
simulation grid (Results), so the numbers below are
illustrative of the design's operating characteristics, not
precision-matched to the main study; a full worked example
with 2,000 replicates is left as future work (see
`docs/pub_review_remediation_2026-08-16.md`).

## Matching Quality

```{r application-balance}
if (!is.null(app_perf)) {
  bal_tab <- app_perf |>
    dplyr::filter(effect == "F1") |>
    dplyr::select(method, mean_max_asmd)
  names(bal_tab) <- c("Method", "Mean max ASMD")
  kable(
    bal_tab,
    booktabs = TRUE,
    digits = 3,
    caption = paste0(
      "Covariate balance at the AHEAD-3-45-like design point ",
      "(N = 500, pool = 2,000, k = 4), by randomization ",
      "method."
    )
  ) |>
    kable_styling(latex_options = c("hold_position"),
                 font_size = 10)
} else {
  cat("*Awaiting `application_example.rds`; run",
      "`analysis/scripts/02_application_example.R`.*")
}
```

## Treatment Effect Estimation

```{r application-estimation}
if (!is.null(app_perf)) {
  est_tab <- app_perf |>
    dplyr::filter(effect %in% c("F1", "F1:F2")) |>
    dplyr::select(method, effect, bias, emp_se, coverage,
                 rejection_rate)
  names(est_tab) <- c("Method", "Effect", "Bias", "Empirical SE",
                      "Coverage", "Rejection rate")
  kable(
    est_tab,
    booktabs = TRUE,
    digits = 3,
    caption = paste0(
      "Bias, empirical SE, 95% CI coverage, and rejection ",
      "rate for the anti-amyloid main effect (F1) and the ",
      "anti-amyloid-by-tau interaction (F1:F2) at the ",
      "AHEAD-3-45-like design point, Neyman estimator."
    )
  ) |>
    kable_styling(latex_options = c("hold_position"),
                 font_size = 9)
} else {
  cat("*Awaiting `application_example.rds`; run",
      "`analysis/scripts/02_application_example.R`.*")
}
```

# Discussion

## Summary of Findings

This manuscript develops matched-tuple randomization for
factorial designs: a greedy tuple-formation algorithm, exact
and conservative variance estimators, a regression-adjusted
estimator, and an asymptotic relative-efficiency result
versus simple randomization. The simulation and worked
example currently on disk (Results; Application) characterize
the operating characteristics of the matched-tuple design
itself, at one and two/four times pool oversampling, across
72 data-generating scenarios plus one AHEAD-3-45-like design
point. As documented in the Provenance note in Results, the
head-to-head empirical comparison against simple, stratified,
and rerandomization baselines -- the paper's central claimed
contribution -- has not yet been regenerated following a bug
fix to the covariate-balance diagnostic, and so is not yet
part of the evidence base for this draft. Until that
comparison exists, the claim that matched-tuple randomization
improves precision relative to those baselines should be read
as a theoretical prediction (Efficiency Analysis) rather than
an established empirical finding.

## Practical Considerations

### When to Use Matched-Tuple Randomization

Matched-tuple randomization is most advantageous when:

1. The full pool of subjects is available before randomization
   (as in trial-ready cohorts), permitting global rather than
   sequential matching.
2. Strong prognostic covariates are measured at baseline.
3. The number of factorial arms $k$ is not too large, to
   avoid excessive degrees-of-freedom loss. The simulation
   and application in this manuscript test only $k = 4$
   ($2^2$ designs); larger $k$ (e.g., $k = 8$ for a $2^3$
   design) is expected to be more susceptible to the
   degrees-of-freedom tradeoff discussed below but has not
   been simulated here, so no specific upper bound on $k$ is
   asserted.
4. The pool-to-sample ratio $r$ exceeds 1, providing excess
   subjects for quality matching.

### Limitations

The primary limitation is the loss of degrees of freedom
relative to unrestricted randomization. With $M$ tuples, the
effective degrees of freedom for variance estimation are
$M - 1$, compared to $N - k$ under simple randomization.
This concern is attenuated when $N$ is large enough that
$M = N/k$ remains sufficient (e.g., $M > 30$).

A second limitation is computational, though of a different
character than originally framed: the package currently ships
only the greedy heuristic (Tuple Formation), which is
$O(N^2)$ in the pairwise distance computation and scales
comfortably to the pool sizes (1,000 to 2,000 characterized
subjects) typical of the TRC setting. An exact
integer-programming formulation, which would guarantee a
globally optimal partition rather than the greedy
approximation, is a natural extension but has not been
implemented or benchmarked in this repository; its
feasibility for $N$ in the low thousands is plausible given
published assignment-problem solve times but is not verified
here.

### Relation to Rerandomization

Rerandomization [@morganRerandomization2012] and matched-tuple
randomization address the same goal -- covariate balance --
through different mechanisms. Rerandomization restricts the
allocation space to a subset of all possible randomizations.
Matched-tuple randomization restricts both the allocation
space and the subject grouping. The two approaches can be
combined: form tuples, then rerandomize within tuples to
further balance residual covariates, though the marginal
benefit of this combination may be small when tuples are
well matched.

## Future Work

Several extensions merit further investigation:

- **Fractional factorials.** When $2^F$ is too large, a
  $2^{F-p}$ fractional factorial reduces $k$. Matched-tuple
  randomization extends directly with $k = 2^{F-p}$.
- **Unequal allocation.** When factorial arms have unequal
  target sizes (e.g., shared placebo arm), the matching
  problem requires tuples of heterogeneous size.
- **Adaptive enrichment.** Combining matched-tuple
  randomization with biomarker-driven adaptive enrichment
  designs.
- **Time-to-event outcomes.** Extending the inference
  framework to survival endpoints.

# References

## Data and Code Availability

```{r sys-info, include=FALSE}
r_version <- paste(R.version$major, R.version$minor, sep = ".")
pkg_version <- tryCatch(
  as.character(utils::packageVersion("factorialrandom")),
  error = function(e) "unknown"
)
mass_version <- tryCatch(
  as.character(utils::packageVersion("MASS")),
  error = function(e) "unknown"
)
sys_name <- Sys.info()[["sysname"]]
```

All source code (R package `factorialrandom`, simulation
drivers, and this manuscript) is version-controlled in the
repository at `~/prj/res/16-factorial-matched-random`
(inner package repository `factorialrandom`). The main
simulation grid is produced by
`analysis/scripts/01_run_simulation.R` and saved to
`analysis/data/derived_data/sim_performance.rds`; the
Application section's worked example is produced by
`analysis/scripts/02_application_example.R` and saved to
`analysis/data/derived_data/application_example.rds`. No
external or restricted-access data are used; all covariate
and outcome data are simulated. This document was rendered
with R `r r_version` on `r sys_name`, package
`factorialrandom` version `r pkg_version`, and `MASS` version
`r mass_version`. Wall-clock time for the full simulation
sweep (5 methods x 72 scenarios x 2,000 replicates) has not
been recorded for the corrected code (see the Provenance note
in Results); a reduced-scale timing check on a single core
(20 replicates x 4 scenarios x 5 methods) took approximately
13 seconds, i.e., roughly 32 ms per replicate-method call.

## Morris et al. (2019) ADEMP Compliance

We audited this simulation study against the reporting standards
proposed by Morris, White, and Crowther [-@morris2019using]. The
original audit is at `docs/morris-audit-2026-04-17.md`; the summary
below reflects the current state of `analysis/scripts/01_run_simulation.R`
and `report.Rmd` as of the 2026-08-16 remediation pass (see
`docs/pub_review_remediation_2026-08-16.md`), not the stale April 17
narrative.

**Verdict:** Mostly compliant, with one open gap.

**Resolved since the April 17 audit:**

- `n_replicates = 2000` is now justified via an MCSE target,
  documented in a comment at `01_run_simulation.R:11-15`.
- `RNGkind("L'Ecuyer-CMRG")` is pinned immediately before
  `set.seed()` at `01_run_simulation.R:278-279`.
- The MCSE columns (`mcse_bias`, `mcse_coverage`,
  `mcse_rejection`) are now surfaced in Table
  \@ref(tab:mcse-table) above, rather than left as TODO
  placeholders.

**Still open:**

- No per-replicate `.Random.seed` snapshot is stored, so an
  individual replicate cannot be reproduced in isolation
  (only the full sweep, via the single fixed seed).
- The full simulation sweep still needs to be rerun following
  the `tuple_balance()` and `regression_adjustment()` bug
  fixes described in Results; see the Provenance note above
  for the exact command.
