---
title: "Matched-Tuple Randomization Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matched-Tuple Randomization Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

This vignette walks through the core `factorialrandom` workflow
for a $2 \times 2$ factorial design ($k = 4$): form matched
tuples from baseline covariates, randomize treatment within
tuples, and estimate factorial effects.

```{r setup}
library(factorialrandom)
set.seed(20260816)
```

## 1. Simulate baseline covariates

In practice, `dat` would be a data.frame of characterized
trial-ready-cohort subjects. Here we simulate a small pool.

```{r simulate-data}
n_pool <- 80
dat <- data.frame(
  X1 = rnorm(n_pool),
  X2 = rnorm(n_pool),
  X3 = rnorm(n_pool)
)
head(dat)
```

## 2. Form matched tuples

`match_tuples()` partitions the pool into tuples of size `k`
(here `k = 4`, for a $2 \times 2$ design) using Mahalanobis
distance by default.

```{r match-tuples}
tuples <- match_tuples(
  dat, k = 4L,
  covariates = c('X1', 'X2', 'X3'),
  distance = 'mahalanobis',
  method = 'greedy'
)
tuples$n_tuples
length(tuples$unmatched)
```

## 3. Randomize within tuples

Each tuple's $k$ members are randomly assigned the $k$ distinct
treatment combinations.

```{r randomize}
treatment <- randomize_tuples(tuples)
table(treatment, useNA = 'ifany')
```

## 4. Check covariate balance

`tuple_balance()` reports the maximum and mean absolute
standardized mean difference (ASMD) across covariates and arms,
for both matched designs and (with `tuples = NULL`) unmatched
designs such as simple randomization.

```{r balance}
bal <- tuple_balance(dat, tuples, c('X1', 'X2', 'X3'), treatment)
bal
```

## 5. Simulate outcomes and estimate effects

`generate_factorial_trial()` wraps the above steps and adds
simulated outcomes; `estimate_effects()` computes main-effect and
interaction estimates with matched-tuple-appropriate standard
errors.

```{r estimate}
trial <- generate_factorial_trial(
  n = 80, n_factors = 2L, p_covariates = 3L,
  r_squared = 0.3, tau_main = c(0.2, 0.2),
  tau_interaction = 0.15, sigma_y = 1, cov_rho = 0.3,
  randomization = 'matched', pool_ratio = 1.0
)

eff <- estimate_effects(
  outcome = trial$data$y,
  treatment = trial$treatment,
  tuples = trial$tuples,
  n_factors = 2L,
  method = 'neyman'
)
eff
```

## 6. Unmatched designs

For comparator designs (simple, stratified, rerandomization),
`generate_factorial_trial()` returns `tuples = NULL`, and both
`tuple_balance()` and `estimate_effects()` accept this directly.

```{r unmatched}
trial_simple <- generate_factorial_trial(
  n = 80, n_factors = 2L, p_covariates = 3L,
  randomization = 'simple'
)
is.null(trial_simple$tuples)

eff_simple <- estimate_effects(
  outcome = trial_simple$data$y,
  treatment = trial_simple$treatment,
  tuples = trial_simple$tuples,
  n_factors = 2L,
  method = 'neyman'
)
eff_simple
```
